---
title: "bha_ls"
---
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(RcappeR)
```

Return lengths per second scale employed by the BHA when they calculate lengths beaten.  Use it to convert the lengths back into seconds, see also [bhascale](bhascale.html)

### Usage
```{r eval=FALSE}
bha_ls(conditions = NULL)
```

### Arguments

param | details
------|-------
conditions | abbreviated concatenation of race conditions, ie. the type of race (_flat_ or _national hunt_) and the official going.

### Conditions

To view the scales and corresponding abbreviations (for the **conditions** param) leave the function blank, ie.

```{r eval=FALSE}
bha_ls()
```

race type | going | scale | conditions
----------|-------|-------|-----------
flat | good or quicker | 6.0 | f-gq
flat | good, good to soft | 5.5 | f-ggs
flat | good to soft, good | 5.5 | f-gsg
flat | good to soft | 5.5 | f-gs
flat | good to soft, soft | 5.5 | f-gss
flat | soft, good to soft | 5.0 | f-sgs
flat | soft or slower | 5.0 | f-ss
flat | all weather | 6.0 | f-aw
flat | fibresand | 5.0 | f-fs
national hunt | good or quicker | 5.0 | nh-gq
national hunt | good, good to soft | 4.5 | nh-ggs
national hunt | good to soft, good | 4.5 | nh-gsg
national hunt | good to soft | 4.5 | nh-gs
national hunt | good to soft, soft | 4.5 | nh-gss
national hunt | soft, good to soft | 4.0 | nh-sgs
national hunt | soft or slower | 4.0 | nh-ss
national hunt | all weather | 5.0 | nh-aw
national hunt | fibresand | 4.0 | nh-fs

[Link to PDF](http://www.britishhorseracing.com/wp-content/uploads/2014/04/Lengths-Per-Second-Scale-Tables.pdf) with the BHA lengths per second scale.

### Example Use

A "National Hunt" race run on "good to soft, soft" ground

```{r}
bha_ls(conditions = "nh-gss")
```
