---
title: "conv_margins"
---
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(RcappeR)
```

Converts margins, between the winner of a race and the horses behind, from lengths into seconds

### Usage
```{r eval=FALSE}
conv_margins(btn_l, cum_l = TRUE, win_time = NULL, conditions = NULL)
```

### Arguments

param | details
------|-------
btn\_l | beaten lengths (numeric)
cum\_l | indicates whether btn\_l is cumulative, default TRUE
win\_time | winners time, optional param, if entered the function returns individual times for each runner
conditions | race conditions that influence the BHA lengths per second scale to be used

### Details

As default, the lengths per second scale used is the US convention of 5 lengths per second.  However, if the race was run under the jurisdiciton of the BHA, then entering an abbreviation of the races' conditions will apply the lengths per second scale employed by the BHA.  See also [bha_ls](bha_ls.html) and [bhascale](bhascale.html)

### Example Use

```{r}
btn_l <- c(0, 0.2, 0.5, 1, 1.5)

conv_margins(btn_l = btn_l)

conv_margins(btn_l = btn_l, cum_l = FALSE)

conv_margins(btn_l = btn_l, win_time = 59)

conv_margins(btn_l = btn_l, win_time = 59, conditions = "f-gq")
```
