---
title: "conv_times"
---
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(RcappeR)
```
Most Horse Racing times are recorded as 'minutes:seconds:milliseconds', data will typically be in character format, the `conv_times` function converts these times into a time in 'seconds.milliseconds'.

### Usage
```{r eval=FALSE}
conv_times(times, regex = NULL)
```

### Arguments

param | details
------|-------
times | times (in character format) to be converted
regex | regular expression to split times by

### Details

Takes a character string representing a race time, splits it up according to the regular expression `[:punct:]]\\s?|\\s?[[:alpha:]]+\\s?|\\s+`, so times like **1m39.99**, **1 39.99**, **1 min 39.99** or **1:39.99** are split up into **1**, **39** and **99** (ie. minutes, seconds, milliseconds).  The function the converts these into seconds and sums them up, returning a numeric time in seconds.  The `regex` paramater allows for custom regular expressions to be entered.

### Example Use

```{r}
times <- c("1.39.99", "1m39.99", "1m 39.99", "1:39.99", "1 min 39.99")
conv_times(times = times)
```
