---
title: "data_prep"
---

```{r, echo=FALSE}
library(RcappeR)
```

Simple, but often necessary, cleaning has been covered in the [**Data Cleaning**](data_cleaning.html) vignette, this vignette will look to more tools to help prepare a dataset for handicapping, as well as some preliminary analysis.

## gulfstream Dataset

The dataset used is included in the package, called **gulfstream**, and contains a good number of races run over the 6 and 8 furlong dirt course at Gulfstream Park in the US.  To load the dataset, and inspect its structure (see `?gulfstream` for details about the variables):

```{r}
data(gulfstream)
str(gulfstream)
```

The dataset is already quite clean, but there are a few functions available in the `RcappeR` package to work with the data.  First a look at the necessary steps ahead of handicapping using either `zipf_init` or `zipf_hcp`.

# Handicapping preparation

In order to use some of the more complex functions (`zipf_init`, `zipf_hcp`) a certain amount of preparation is required.  There are a number of variables needed for handicapping, these are:

* unique race id (a way to distinguish different races)
* race classes, or types
* surface
* distance
* final times
    + either times for all runners, or
    + a winning time and a way to calculate final times (beaten seconds, or beaten lengths)
* weight carried

The variables above should be pretty common in a racing dataset that you wish to calculate ratings from.  In the **gulfstream** dataset we have all the above.  Individual final times for horses might be a hurdle, but lengths beaten is a much more common variable, and as covered in the [**Data Cleaning**](data_cleaning.html) vignette, the `conv_margins` can convert lengths beaten into final times.

A unique race id is required in the **gulfstream** dataset, but this can be created by concatenating the `date` and `race` variables.  Obviously if a dataset contains races at more than one racecourse, it would be wise to include something about that: you can't have two races being run at the same track, on the same day at the same time.  Let's create a variable called `date_race`:

```{r}
gulfstream$date_race <- paste(gulfstream$date, gulfstream$race, sep = "_")
```

The above `date_race` variable was the only one missing from the above list, but before handicapping can begin we need to calculate margins between horses that take into account the following:

* distance
* surface
* beaten margins
* weight carried

The best (imo) way to do this is to use the package `dplyr` which takes advantage of the `%>%` pipe function from `magrittr` to calculate the necessary variables.  The code below processes the gulfstream dataset, creating the necessary variables.  It is explained in more detail below the code, the functions used from `RcappeR` are `btn_sec`, `lbs_per_sec` and `diff_at_wgts`:

```{r warning=FALSE, message=FALSE}
library(dplyr)
gulfstream <- gulfstream %>%
    group_by(date_race) %>%
    mutate(btn_sec = btn_sec(fintime),
           scale = lbs_per_sec(dist = dist, surf = "dirt"),
           btn_lbs = scale * btn_sec,
           diff_wgts = diff_at_wgts(btn_lbs = btn_lbs, wgt_carried = wgt))
```

1. Load dplyr library
    + `library(dplyr)`
2. Use gulfstream dataset and store results back in gulfstream
    + `gulfstream <- gulfstream %>%`
3. Group races by the unique race ids (date_race)
    + `group_by(date_race)`
4. Create new variables within each group (each race)
    + calculate margins (in seconds) between horses
        + `btn_sec = btn_sec(fintime),`
    + calculate lbs per second scale
        + `scale = lbs_per_sec(dist = dist, surf = "dirt"),`
    + calculate beaten lbs
        + `btn_lbs = scale * btn_sec,`
    + calculate difference at the weights (margins, and weight carried)
        + `diff_at_wtgs(btn_lbs = btn_lbs, wgt_carried = wgt))`
        
At this stage handicapping can begin by using the `zipf_init` function, see [**Initialising a Handicap**](initialise_handicap.html) vignette.  Ahead of using `zipf_race` and `zipf_hcp` users will perform the same steps as above, but with a dataset containing a single race (so no need for _group\_by_) that needs to be handicapped, the code shows the theoretical steps for one race.

```{r eval=FALSE}
one_race <- one_race %>%
    mutate(btn_sec = btn_sec(fintime),
           scale = lbs_per_sec(dist = dist, surf = "dirt"),
           btn_lbs = scale * btn_sec,
           diff_wgts = diff_at_wgts(btn_lbs = btn_lbs, wgt_carried = wgt))
```

# Preliminary Analysis

## Beaten margins

Unlike the dataset used in the **Data Cleaning** vignette we have individual times for all runners, but no margins in seconds.  The `btn_sec` function can calculate margins between horses, necessary when handicapping races.  If a dataset only has one race then it can be used as such:

```{r eval=FALSE}
btn_sec(times = fintimes)
```

However, as the **gulfstream** dataset contains many different races, the `date_race` is needed to identify the unique races, otherwise `btn_sec` would calculate the margins between the fastest time in the dataset and each of the `r length(gulfstream$date)` rows.  Margins between the leader after 4 furlongs can also be calculated.

```{r}
gulfstream <- gulfstream %>%
    group_by(date_race) %>%
    mutate(btn_sec = btn_sec(times = fintime),   ## behind winner at the finish
           bhd_ldr_sect = btn_sec(times = sect_4f))  ## behind leader at the 4 furlong sectional
```

## Finishing Speeds

Finishing Speed is an innovation by Simon Rowlands of Timeform, find more details in posts by Simon, [Part 1](https://www.timeform.com/Racing/Articles/An_introduction_to_Sectional_Timing_Part_1) and [Part 2](https://www.timeform.com/Racing/Articles/An_introduction_to_Sectional_Timing_Part_2).  In short it compares the finishing speed of a horse to its average race speed; a finishing speed of 100 means the horse showed the same speed over the finishing sectional as it did the entire race; lower than 100 and the horse finished slower; higher, faster.

In the dataset there are individual times for all horses after 4 furlongs in their race, variable `sect_4f`, and their final time, `fintime`.  The `fin_spd` function needs the time for the closing sectional, as well as the closing sectional distance - which is the race distance minus 4 (furlongs).

```{r}
# calculate the closing sectional time
gulfstream$sect_time <- gulfstream$fintime - gulfstream$sect_4f
# using the new sect_time variable and other data in the dataset
gulfstream$fin_spd <- fin_spd(fin_time = gulfstream$fintime, 
                              dist = gulfstream$dist, 
                              sect_time = gulfstream$sect_time, 
                              sect_dist = gulfstream$dist - 4)
```

The following plot shows our newly created finishing speed variable plotted on the y axis against the times horses recorded over the opening 4f.  Horses over the 8f dirt course finish their closing sectional (4 furlongs) faster than horses running the 6f dirt course do their closing sectional (2 furlongs).

```{r, fig.align='center', echo=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(ggplot2))
ggplot(gulfstream, aes(x = sect_4f, y = fin_spd)) +
    geom_point(alpha=.5, color = "#D8D8D8") +
    geom_point(data = subset(gulfstream, pos == 1), aes(x = sect_4f, y = fin_spd), col="#D9220F", alpha = .65) +
    theme_rcapper() +
    facet_wrap(~dist)
```

Using the `dplyr` package it is easy to calculate par Finishing Speeds for each finishing position, from which further analysis can be carried out (though it won't be here).  Winners, unsurprisingly, have higher finishing speeds than those in behind.  Focusing on just 6 furlong races

```{r, warning=FALSE, message=FALSE}
gulfstream %>%
    filter(dist == 6) %>%
    group_by(pos) %>%
    summarise(par_fs = mean(fin_spd, na.rm = TRUE),
              sd_fs = sd(fin_spd, na.rm = TRUE))
```

## Runners Beaten

Draw analysis is often a quick and dirty way of eliminating runners, but analysis often just looks at raw counts, winners vs loses.  The `rnrs_btn` function calculates the percentage of runners beaten, which obviously benefits from looking at **ALL** runners in a race, rather than just the winners.  Once again we need the `date_race` variable and the `dplyr` package to isolate individual races.

```{r}
gulfstream %>%
    group_by(date_race) %>%
    mutate(pct_btn = rnrs_btn(pos = pos)) %>%
    ungroup() %>%
    group_by(gate) %>%
    summarise(pct_btn = mean(pct_btn, na.rm = TRUE))
```