---
title: "fin_spd"
---
```{r echo=FALSE}
library(RcappeR)
```

Calculates finishing speed percentages, an innovation by Simon Rowlands of Timeform.

### Usage
```{r eval=FALSE}
fin_spd(fin_time, dist, sect_time, sect_dist)
```

### Arguments

param | details
------|-------
fin_time | final time
dist | race distance
sect_time | closing sectional time
sect_dist | closing sectional distance

### Details

Parameters must be in the same units, so time params **fin_time** and **sect_time** will be in seconds, while distance params **dist** and **sect_dist** will **both** be in furlongs, feet, metres, miles, etc.

Two articles by Simon Rowlands explaining Finishing Speeds (and sectionals) can be found:
[An Introduction to Sectional Timing (Part 1)](https://www.timeform.com/Racing/Articles/An_introduction_to_Sectional_Timing_Part_1)
[An Introduction to Sectional Timing (Part 2)](https://www.timeform.com/Racing/Articles/An_introduction_to_Sectional_Timing_Part_2)

### Example Use

We'll calculate the finishing speed for a horse running in a 5 furlong race, the horse had a final time of 59 seconds, and ran the closing 2 furlongs in 23 seconds

```{r}
fin_spd(fin_time = 59, dist = 5, sect_time = 23, sect_dist = 2)
```