---
title: "Initialising a Handicap"
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(RcappeR)
```

This vignette walks through the use of `zipf_init`, which initialises a handicap using a collection of races, in this case the **gulfstream** dataset (used in [**Data Preparation**](data_prep.html) vignettes).  There are certain steps necessary ahead of using `zipf_init`, these are covered in the **Data Preparation** vignette, but will be covered again here.

## gulfstream Dataset

Load the dataset:
```{r}
data(gulfstream)
```

The **gulfstream** dataset contains `r length(unique(paste(gulfstream$date, gulfstream$race, sep = "_")))` unique races, this is not a huge number, the more races the better.  A look at the structure of **gulfstream**.

```{r}
str(gulfstream)
```

# Handicapping preparation

In order to use some of the more complex functions (`zipf_init`, `zipf_hcp`) a certain amount of preparation is required.  There are a number of variables needed for handicapping, these are:

* unique race id (a way to distinguish different races)
* race classes, or types
* surface
* distance
* final times
    + either times for all runners, or
    + a winning time and a way to calculate final times (beaten seconds, or beaten lengths)
* weight carried

The variables above should be pretty common in a racing dataset that you wish to calculate ratings from.  In the **gulfstream** dataset we have all the above.  Individual final times for horses might be a hurdle, but lengths beaten is a much more common variable, and as covered in the [**Data Cleaning**](data_cleaning.html) vignette, the `conv_margins` can convert lengths beaten into final times.

A unique race id is required in the **gulfstream** dataset, but this can be created by concatenating the `date` and `race` variables.  Obviously if a dataset contains races at more than one racecourse, it would be wise to include something about that: you can't have two races being run at the same track, on the same day at the same time.  Let's create a variable called `date_race`:

```{r}
gulfstream$date_race <- paste(gulfstream$date, gulfstream$race, sep = "_")
```

The above `date_race` variable was the only one missing from the above list, but before handicapping can begin we need to calculate margins between horses that take into account the following:

* distance
* surface
* beaten margins
* weight carried

The best (imo) way to do this is to use the package `dplyr` which takes advantage of the `%>%` pipe function from `magrittr` to calculate the necessary variables.  The code below processes the gulfstream dataset, creating the necessary variables.  It is explained in more detail below the code, the functions used from **RcappeR** are `btn_sec`, `lbs_per_sec` and `diff_at_wgts`:

```{r warning=FALSE, message=FALSE}
library(dplyr)
new_gulfstream <- gulfstream %>%
    group_by(date_race) %>%
    mutate(btn_sec = btn_sec(fintime),
           scale = lbs_per_sec(dist = dist, surf = "dirt"),
           btn_lbs = scale * btn_sec,
           diff_wgts = diff_at_wgts(btn_lbs = btn_lbs, wgt_carried = wgt))
```

1. Load dplyr library
    + `library(dplyr)`
2. Use gulfstream dataset and store results back in gulfstream
    + `gulfstream <- gulfstream %>%`
3. Group races by the unique race ids (date_race)
    + `group_by(date_race)`
4. Create new variables within each group (each race)
    + calculate margins (in seconds) between horses
        + `btn_sec = btn_sec(fintime),`
    + calculate lbs per second scale
        + `scale = lbs_per_sec(dist = dist, surf = "dirt"),`
    + calculate beaten lbs
        + `btn_lbs = scale * btn_sec,`
    + calculate difference at the weights (margins, and weight carried)
        + `diff_at_wtgs(btn_lbs = btn_lbs, wgt_carried = wgt))`

At this stage, the **gulfstream** dataset can be entered into `zipf_init`.  First a word about the methodology for initialising the handicap.

## Handicapping methodology

The handicapping methodology uses a version of race standardisation first explained by Simon Rowlands, Head of Research at Timeform, specifically using Zipfs Law (hence the names of this family of functions, see also `?zipf_race` and `?zipf_hcp`).

Race standardisation looks at races of similar class/type and assesses the performance of one winner, by assessing the performance of winners in the different, but similar, races.  A more detailed explanation can be found in the [**Handicap with zipf_race**](handicap_with_zipf_race.html) vignette, which walks through a simple example using the `zipf_race` function, which is called by `zipf_init` (and `zipf_hcp`).

## Initialising a handicap - zipf_init

Race standardisation uses past ratings from similar types/classes of race to assess a new race, in initialising a handicap there are no past ratings.  So the `zipf_init` function group races together and assess performances using margins between horses - the `diff_wgts` variable created above.  This process builds a skeleton handicap, from which further handicapping can, and should, be undertaken.

Below is a simple table explaining the various inputs to `zipf_init`:

param | details | example input
------|---------|---------
races | a dataframe of races | `new_gulfstream`
group\_by | name(s) of variables to group races by | `"race_type"` (could also include `value`)
race\_id | name of variable to identify the unique races in the `races` dataframe | `"date_race"`
btn\_var | name of variable containing margins between horses in `races` dataframe | `"diff_wgts"`
.progress | plyr's progress bar, useful when using on large datasets (>20k rows) as the function takes time to run | "text"

So:
```{r echo=FALSE}
start.time <- Sys.time()

our_hcp <- zipf_init(races = new_gulfstream, group_by = "race_type", race_id = "date_race", btn_var = "diff_wgts")

end.time <- Sys.time()
time.taken <- end.time - start.time
```

```{r eval=FALSE}
our_hcp <- zipf_init(races = new_gulfstream, group_by = "race_type", race_id = "date_race", btn_var = "diff_wgts", .progress = "text")
```

This small example, handicapping `r length(unique(new_gulfstream$date_race))` races, split into `r length(unique(new_gulfstream$race_type))` different race types (`r unique(new_gulfstream$race_type)`), took `r time.taken` seconds.

The output from `zipf_init` is a list (of class "rcapper\_zipf\_init"), there are print and summary methods for this class of object (though both do the same):

```{r}
our_hcp

summary(our_hcp)
```

There is also a plot method, perhaps the most useful, which plots the distribution of ratings for each group, as we can see below the small samples in a couple of the race types shows the need for more races, or at least making sure groups are of a decent size.

```{r fig.align='center'}
plot(our_hcp)
```

The plot shows a distribution of ratings (in lbs) for the winners in the `r length(unique(new_gulfstream$date_race))` races in `new_gulfstream` dataset.  The mean will always be around 0, for all race types.  The next step is to assign a standard rating for a winner of this type/class of race.  These standards should reflect the difference in ability (in lbs) between the different race types, so a standard rating for Grade 1 winner is going to be far greater than that of a Maiden race, what these differences are is unknown - I am working on a solution to help find these differences.

Possible solutions to this issue is to use ratings from other handicappers to help guide this process, for example, Timeform (including Timeform US) or Beyer class pars.

Finally, `merge_zipf_init` function will merge the resulting ratings from `zipf_init` with the dataset used to calculate the ratings.  Finally print the first 20 rows, showing the variables created in this vignette and the zipf_rtg for runners:

```{r}
initial_hcp <- merge_zipf_init(zipf_list = our_hcp, races = new_gulfstream, btn_var = "diff_wgts")
# Let's have a look at the first few rows of our skeleton handicap
initial_hcp %>%
    select(race_type, date_race, pos, fintime, btn_sec:zipf_rtg) %>%
    head(15)
```


<a href="https://github.com/durtal/RcappeR" class="github-corner"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#d9230f; color:#fcfcfc; position: absolute; top: 0; border: 0; right: 0;"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>
