---
title: "lbs_per_sec"
---
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(RcappeR)
```

Returns a lbs per second scale taking into account race distance and surface

### Usage
```{r eval=FALSE}
lbs_per_sec(dist, surf = "turf")
```

### Arguments

param | details
------|-------
dist | race distance (in furlongs, only races between 5 and 16)
surf | race surface, default is "turf", other option is "dirt" (see details)

### Details

Uses a loess model to return a simple lbs per second scale.  The **dist** parameter must be between 5 and 16, values outside this range will return NA.  If the surface is fibresand, then use the "dirt" option for the **surf** parameter, if all-weather then use the "turf" option.  

Simple model comes from Bob Wilkins' book [Bioenergetics and Racehorse Ratings](http://www.amazon.co.uk/Bioenergetics-Racehorse-Ratings-Bob-Wilkins/dp/0956424309).  This function should eventually be replaced by a more complex/flexible model that takes into account the speed a horse ran its race.

### Example Use


```{r}
lbs_per_sec(dist = 5, surf = "turf")

lbs_per_sec(dist = 5, surf = "dirt")
```
