---
title: "zipf_hcp"
---
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(RcappeR)
```

Assess the performance (ie. handicap) of a winner of a race using race standardisation; which uses the performances of runners in a different, but similar, race.  This function calls [zipf_race](zipf_race.html), and uses many races to handicap the winner of a single race.  See the [Handicapping using zipf_race](handicap_with_zipf_race.html) vignette for a detailed explanation of the use of Zipfs Law, and also [Handicapping using zipf_hcp](handicap_with_zipf_hcp.html) to see **zipf\_hcp** in use.

### Usage
```{r eval=FALSE}
zipf_hcp(race, past_races, race_id, btn_var, rating = NULL, results = "detail", .progress = "none")
```

### Arguments

param | details
------|-------
race | dataframe of a race to handicap
past\_races | dataframe of past races to be used to handicap **race**
race\_id | name of variable to split **past\_races** up by so each split is one race
btn\_var | name of variable in **race** that contains the margins (in lbs) between the horses
rating | name of ratings variable (if applicable) in race\_2
results | default "detail", determines the output, other option is "simple", which will return the mean rating of all possible ratings
.progress | **plyr**'s progress bar, default is "none", options inc. "text", "time", "tk" or "win"

### Details

The **past\_races** dataframe is split according to **race\_id**, so each split should be a small dataframe of a single race.  For each of these single race dataframes, they are used as the **race\_2** parameter in [zipf_race](zipf_race.html), while the race being handicapped is used as the **race** parameter.  

If _simple_ is entered into the **results** parameter then a single rating, the mean of all possible ratings, is returned.  If the default of _detail_ is left, then a list is returned containing:

* **n\_races** the number of unique races used to handicap **race**
* **mean\_rtg** mean rating for **race** (this is the value returned when _simple_ is entered in **results**)
* **summary** summary statistics of ratings returned
* **ratings** dataframe of ratings (with variables _race\_id_ and _zipf\_rtg_)

An article by Simon Rowlands explaining his use of Zipfs Law and race standardisation can be found [here](https://betting.betfair.com/horse-racing/bloggers/simon-rowlands/simon-rowlands-on-handicapping-060710.html)

<a href="https://github.com/durtal/RcappeR" class="github-corner"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#d9230f; color:#fcfcfc; position: absolute; top: 0; border: 0; right: 0;"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>
